pirouette: the error BEAST2 makes in inferring a phylogeny

### 2019-01-29 article.tex
Use '\&' instead of '$\&$'

Phylogenies are commonly used to explore evolutionary hypotheses.
The mode of diversification is one such hypothesis.
Phylogenies constructed from DNA alignments are commonly fit to
diversification models, or the other way around. For example,
[Etienne $\&$ Rosindell 2012] fit four bird phylogenies to the protracted
diversification model.
\giovanni{ALTERNATIVE VERSION: The development of new powerful inference tools, such as BEAST [\cite{drummond2007beast}] or MrBayes [\cite{huelsenbeck2001mrbayes}], provided us the possibility to build phylogenetic trees from genetic material extracted from extant organisms. This has constituted a big step forward in the general understanding on how species evolved (this can be improved). Having access to this new kind of data has given rise to a high interest in understanding what are the main drivers and modes for diversification processes. }

There are plenty of diversification models: constant-birth [Yule, 19..],
birth-death [\cite{nee1994reconstructed}], time-dependent [?], diversity 
dependent [\cite{etienne2011diversity}], protracted 
birth-death [\cite{rosindell2010protracted}][\cite{etienne2012prolonging}],
multiple-birth death [Laudanno, Bilderbeek, Haegeman $\&$ Etienne, in preparation],
a rate shifting birth-death model [the SLS model. Laudanno, Haegeman $\&$ Etienne, in preparation]
punctuated equilibrium aka early burst [?Harmon],
dependent on a binary trait [\cite{maddison2007estimating}], 
multiple state trait [\cite{fitzjohn2012diversitree}],
concealed state [\cite{beaulieu2016detecting}] or even a combination of multiple concealed and observable states [\cite{herrera2018detecting}].
\giovanni{
@Richel: I am wondering if it might be worth to give a brief explanation on the philosophy of the aforementioned models, to explain why we want to simulate the processes. In case, this could be a stub for it:
Such models usually rely on the assumption that a particular diversification process can be explained mainly by specific mechanisms.
Such mechanisms are built into likelihood functions, which provide the likelihood of the parameters, given the data.
By maximizing the likelihood is possible to give the best inference of such parameters. It is therefore a standard practice, to validate such models, to test them versus simulated processes generated according to the hypothesised mechanisms.
}
Diversification models are getting more complex over time. There are many
reasons for this. One reason is the increase in computing power,
which allows this heightened complexity possible.
Another reason is to explore a process that is not
yet captured by the other diversification models. An example of this
is the multiple-birth model [Laudanno, Bilderbeek, Haegeman $\&$ Etienne, in preparation], 
that is the first model to allow speciation to co-occur.

@@

\verb;pirouette; is an R package that performs such an experiment,
and is built on \verb;babette; [Bilderbeek $\&$ Etienne, 2018], 
which calls the popular Bayesian inference tool 
BEAST2 [\cite{bouckaert2014beast}]. With \verb;pirouette;, one
can easily measure the error made by Bayesian inference in recovering
any given phylogeny, helping us to valorize new diversification models.

@@

The fourth step measures the inference error, using the phylogenies in the
Bayesian posterior. These phylogenies are compared to the given
phylogeny using an error statistic, which is the nLTT 
statistic (\cite{janzen2015approximate}) by default,
but also a custom error statistic is supplied, based
on the absolute difference in the gamma statistic [Pybus $\&$ Harvy]. 
Additionally, the user can specify the
proportion of posterior phylogenies to 
discard (i.e. the burn-in), throwing away the first $10\%$
of all phylogenies by default. This burn-in is used to discard
the part of the MCMC chain that has not yet converged to a
representative part of state space.

>>>>

Phylogenies are commonly used to explore evolutionary hypotheses.
The mode of diversification is one such hypothesis.
Phylogenies constructed from DNA alignments are commonly fit to
diversification models, or the other way around. For example,
[Etienne \& Rosindell 2012] fit four bird phylogenies to the protracted
diversification model.
\giovanni{ALTERNATIVE VERSION: The development of new powerful inference tools, such as BEAST [\cite{drummond2007beast}] or MrBayes [\cite{huelsenbeck2001mrbayes}], provided us the possibility to build phylogenetic trees from genetic material extracted from extant organisms. This has constituted a big step forward in the general understanding on how species evolved (this can be improved). Having access to this new kind of data has given rise to a high interest in understanding what are the main drivers and modes for diversification processes. }

There are plenty of diversification models: constant-birth [Yule, 19..],
birth-death [\cite{nee1994reconstructed}], time-dependent [?], diversity 
dependent [\cite{etienne2011diversity}], protracted 
birth-death [\cite{rosindell2010protracted}][\cite{etienne2012prolonging}],
multiple-birth death [Laudanno, Bilderbeek, Haegeman \& Etienne, in preparation],
a rate shifting birth-death model [the SLS model. Laudanno, Haegeman \& Etienne, in preparation]
punctuated equilibrium aka early burst [?Harmon],
dependent on a binary trait [\cite{maddison2007estimating}], 
multiple state trait [\cite{fitzjohn2012diversitree}],
concealed state [\cite{beaulieu2016detecting}] or even a combination of multiple concealed and observable states [\cite{herrera2018detecting}].
\giovanni{
@Richel: I am wondering if it might be worth to give a brief explanation on the philosophy of the aforementioned models, to explain why we want to simulate the processes. In case, this could be a stub for it:
Such models usually rely on the assumption that a particular diversification process can be explained mainly by specific mechanisms.
Such mechanisms are built into likelihood functions, which provide the likelihood of the parameters, given the data.
By maximizing the likelihood is possible to give the best inference of such parameters. It is therefore a standard practice, to validate such models, to test them versus simulated processes generated according to the hypothesised mechanisms.
}
Diversification models are getting more complex over time. There are many
reasons for this. One reason is the increase in computing power,
which allows this heightened complexity possible.
Another reason is to explore a process that is not
yet captured by the other diversification models. An example of this
is the multiple-birth model [Laudanno, Bilderbeek, Haegeman \& Etienne, in preparation], 
that is the first model to allow speciation to co-occur.

@@

\verb;pirouette; is an R package that performs such an experiment,
and is built on \verb;babette; [Bilderbeek \& Etienne, 2018], 
which calls the popular Bayesian inference tool 
BEAST2 [\cite{bouckaert2014beast}]. With \verb;pirouette;, one
can easily measure the error made by Bayesian inference in recovering
any given phylogeny, helping us to valorize new diversification models.

@@

The fourth step measures the inference error, using the phylogenies in the
Bayesian posterior. These phylogenies are compared to the given
phylogeny using an error statistic, which is the nLTT 
statistic (\cite{janzen2015approximate}) by default,
but also a custom error statistic is supplied, based
on the absolute difference in the gamma statistic [Pybus \& Harvy]. 
Additionally, the user can specify the
proportion of posterior phylogenies to 
discard (i.e. the burn-in), throwing away the first $10\%$
of all phylogenies by default. This burn-in is used to discard
the part of the MCMC chain that has not yet converged to a
representative part of state space.
